---
title: "Expository Graphs"
subtitle: "JHU Data Science"
author: "www.jtleek.com/advdatasci"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "../additional.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    keep_md: true
---
class: inverse, middle, center


# Expository Graphs: tell a story


```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE,
  prompt = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  results = 'hide',
  fig.height = 5.5,
  fig.width = 10
  )
```

---
class: inverse, middle, center


# In papers, figures (+ caption) should be <br><br> .toobig[self-sufficient]

---
class: inverse

## A first plot - not polished

```{r plot}
library(ggplot2)
g = ggplot(
  data = quakes, 
  aes(x = lat,y = long,colour = stations)) + 
  geom_point()
```

.super[
1.  Make a container for data `ggplot`.
2.  Use the `quakes` `data.frame`: `data = quakes`.
3.  Map certain "aesthetics" with the `aes` to three different aesthetics (`x`, `y`, `z`) to certain variables from the dataset `lat`, `long`, `stations`, respectively.
4.  Add a layer of geometric things, in this case points (`geom_point`).
]

---
class: inverse

## Critiques (minimum of 4)

```{r show, echo = FALSE}
print(g)
```


---
class: inverse

## Critiques (at least)

.super[

1. make the axes bigger<br><br>
2. make the labels bigger<br><br>
3. make the labels be full names (latitude and longitude, ideally with units when variables need them<br><br>
4. make the legend title be number of stations reporting<br><br>
]

---
class: inverse

## Theme - get to know it

.super[
- Go to `?theme` (right now on your computer)
<br><br>
- `text` argument to change **ALL** the text sizes to a value. 
<br><br>
- The `text` argument (input) in the `theme` command requires class `element_text`.  
]

---
class: inverse

## Bigger text/labels

```{r bigger_axis}
g + theme(text = element_text(size = 20))
```

---
class: inverse

## Bigger text/labels - but different
```{r bigger_axis2}
tsize = function(size) element_text(size = size)
gbig = g + theme(axis.text = tsize(18), axis.title = tsize(20),
  legend.text = tsize(15), legend.title = tsize(15));
```
```{r bigger_axis2.1, echo = FALSE}
gbig
```



---
class: inverse

## Make the Labels to be full names

```{r lab_full}
gbig = gbig + xlab("Latitude") + ylab("Longitude"); gbig
```


---
class: inverse

## Maybe add a title

```{r title}
gbig + ggtitle("Spatial Distribution of Stations")
```


---
class: inverse

## Bigger title

```{r big_title}
gbig + ggtitle("Spatial Distribution of Stations") + 
  theme(title = element_text(size = 30))
```

---
class: inverse
## Making a better legend

```{r leg}
gbigleg_orig = gbig + guides(colour = guide_colorbar(
  title = "Number of Stations Reporting")); gbigleg_orig
```

---
class: inverse

## Making a better legend: line breaks

```{r leg2}
gbigleg = gbig + guides(colour = guide_colorbar(
  title = "Number\nof\nStations\nReporting")); gbigleg
```

---
class: inverse

## Making a better legend: centered


```{r leg_adjust}
gbigleg = gbigleg + guides(colour = guide_colorbar(
    title = "Number\nof\nStations\nReporting", 
    title.hjust = 0.5));  gbigleg
```


---
class: inverse

## Put legend INSIDE the plot

```{r leg_inside}
gbigleg + theme(legend.position = c(0.3, 0.35))
```

---
class: inverse

## Put legend INSIDE the plot

.super[
Problems:

1.  There may not be enough place to put the legend
2.  The legend may mask out points/data

For problem 1., we can either 1) make the y-axis bigger or the legend smaller or a combination of both.  In this case, we do not have to change the axes, but you can use `ylim` to change the y-axis limits
]

---
class: inverse

## Put legend INSIDE the plot

```{r change_ylim}
gbigleg + theme(legend.position = c(0.3, 0.35)) +
  ylim(c(160, max(quakes$long)))
```


---
class: inverse

## Making a transparent legend

.super[
But that changes the scaling of the plot!

Better solution:
]
```{r trans_leg}
transparent_legend =  theme(
  legend.background = element_rect(fill = "transparent"),
  legend.key = element_rect(fill = "transparent", 
                            color = "transparent")
)
```

---
class: inverse

## Making a transparent legend

```{r leg_inside2}
gtrans_leg = gbigleg + theme(legend.position = c(0.3, 0.35)) +
  transparent_legend; print(gtrans_leg)
```

---
class: inverse

## Comparison again

```{r leg_inside10}
gbigleg + 
  theme(legend.position = c(0.3, 0.35))
```


---
class: inverse

## Moving the title of the legend

```{r leg_left}
gtrans_leg + guides(
  colour = guide_colorbar(title.position = "right"))
```

---
class: inverse

## **Respecifying** guides should be done in one shot:
```{r leg_left_correct}
gtrans_leg + guides(colour = guide_colorbar(
  title = "Number\nof\nStations\nReporting",
  title.hjust = 0.5, title.position = "right"))
```

---
class: inverse

## Changing after the fact

```{r respec}
gtrans_leg$guides$colour$title.position = "left"; gtrans_leg
```


---
class: inverse

## Figure captions: critiques

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

A plot of latitude versus longitude.  

---
class: inverse

## Figure captions: be specific

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

A plot of earthquake latitude versus longitude collected from a cube around Fiji in 1964.  


---
class: inverse

## Figure captions: label the caption

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

Figure 1 A plot of earthquake latitude versus longitude collected from a cube around Fiji in 1964.  

---
class: inverse

## Figure captions: tell a story

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

**Figure 1. Volcanic eruptions occur in clusters.** A plot of earthquake latitude versus longitude collected from a cube around Fiji in 1964.  

---
class: inverse

## Figure captions: include units

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

**Figure 1. Volcanic eruptions occur in clusters.** A plot of earthquake latitude (degrees) versus longitude (degrees) collected from a cube around Fiji in 1964.  

---
class: inverse

## Figure captions: explain aesthetics

```{r, echo = FALSE, fig.height=5.5, fig.width=10}
gtrans_leg
```

**Figure 1. Volcanic eruptions occur in clusters.** A plot of earthquake latitude (degrees) versus longitude (degrees) collected from a cube around Fiji in 1964. Lighter color means more stations reported the earthquake. Darker colors mean fewer stations reporting, and there does not appear to be a strong relationship between geography and the number of stations reporting.  

---
class: inverse

## The full specification

```{r, fig.height=5.5, fig.width=10, fig.show='hide'}
ggplot(
  data = quakes, aes(x = lat,y = long,colour = stations)) + 
  geom_point() + theme(axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15),
    title = element_text(size = 30)) + 
  xlab("Latitude") + ylab("Longitude") + 
  ggtitle("Spatial Distribution of Stations") + 
  guides(colour = guide_colorbar(
    title = "Number\nof\nStations\nReporting",
    title.hjust = 0.5, title.position = "left")) +
  theme(legend.position = c(0.3, 0.35)) +
  transparent_legend
```

---
class: inverse

## "I don't like that theme": `ggtheme` 

```{r themes}
g + theme_bw()
```

---
class: inverse

## "I don't like that theme": `ggtheme` 

```{r theme_dark}
g + theme_dark()
```

---
class: inverse

## "I don't like that theme": `ggtheme` 

```{r theme_minimal}
g + theme_minimal() # no box!
```

---
class: inverse

## "I don't like that theme": `ggtheme` 

```{r theme_classic}
g + theme_classic() # axis lines!
```

---
class: inverse

## "I don't like that theme": `ggthemes` 

```{r theme_base}
g + ggthemes::theme_base()
```

---
class: inverse

## Conclusions


- .super[`ggplot2` can deceive new users by making graphs that look "good"-ish. ]

--

- .super[they are not good enough by default.]

--

- .super[they are not good enough by default.]

--

- .super[they are not good enough by default.]

--

- .super[Doesn't matter `base` vs `ggplot2`]



---
class: inverse

## A typical `ggplot2`: mtcars

Set up a graph (but with no `x` variable)
```{r}
g = ggplot(aes(y = am), data = mtcars) + 
  geom_point(position = position_jitter(height = 0.2)) + 
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), se = FALSE) +
  geom_smooth(method = "loess", se = FALSE, col = "red")
```

.super[
- binary `y` outcome - jittered
- fit a `glm` of the data from the `x`-variable
- fit a `loess` for non-parametric version

]

---
class: inverse

## A typical `ggplot2`: mtcars

```{r}
g + aes(x = mpg)
```


---
class: inverse

## A typical `ggplot2`: mtcars

```{r}
gmpg = g + aes(x = mpg); gdrat = g + aes(x = drat)
gridExtra::grid.arrange(gmpg, gdrat, ncol = 2)
```


---
class: inverse

## A typical `ggplot2`: mtcars

```{r}
gmpg = gmpg + annotate(x = 30, y = 0.2, geom = "text", label = "A", size = 20)
gdrat = gdrat + annotate(x = 4.5, y = 0.2, geom = "text", label = "B", size = 20)
gridExtra::grid.arrange(gmpg,gdrat, ncol = 2)
```


---
class: inverse

## Figure summary

.super[
- Defaults are easily spotted (not always terrible)
<br><br>
- Always label panels
<br><br>
- Always reference panels in figure captions
<br><br>
- If adding a smoother/fit, label it **and reference it**
<br><br>
- Figures should be self-sufficient (first thing people look at)
]

