---
title: "Dimension Reduction"
subtitle: "JHU Data Science"
author: "www.jtleek.com/advdatasci"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "../additional.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    keep_md: true
---
class: inverse, middle, center

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "")
source("../slide_functions.R")
folder = "../imgs/dim_red//"
knitr::opts_chunk$set(
  echo = TRUE,
  prompt = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = ""
  )
library(RColorBrewer)
library(pheatmap)
```

# Dimension Reduction ≈ compression of data

```{r latent, results='asis', echo = FALSE}
bg_slide("pagerank", 
  folder = folder,
  size = "70%",
  add_opts = "class: inverse")
```

.footnote[http://ilpubs.stanford.edu:8090/422/1/1999-66.pdf]


```{r genome_svd, results='asis', echo = FALSE}
bg_slide("genome_svd", 
  folder = folder,
  size = "70%",
  add_opts = "class: inverse")
```

.footnote[http://www.pnas.org/content/97/18/10101.full]

```{r genome_svd2, results='asis', echo = FALSE}
bg_slide("genome_svd2", 
  folder = folder,
  size = "60%",
  add_opts = "class: inverse")
```

.footnote[http://www.nature.com/ng/journal/v38/n8/full/ng1847.html]

---
class: inverse

## Matrix Data

```{r pars, echo = FALSE}
opar = par(no.readonly = TRUE)
set.seed(12345)
par(mar = rep(0.2, 4))
```

```{r plot, fig.width = 8, fig.height = 6}
dataMatrix <- matrix(rnorm(400), nrow = 40)
image(1:10, 1:40, t(dataMatrix)[, nrow(dataMatrix):1])
```

---
class: inverse
## Matrix Data

```{r pheatmap, fig.width = 8, fig.height = 6}
pheatmap::pheatmap(dataMatrix)
```


```{r pars_reset, echo = FALSE}
par(opar)
```

---
class: inverse
## Matrix Data: add a pattern

```{r pattern}
set.seed(678910)
for(i in 1:40){
  # flip a coin
  coinFlip <- rbinom(1,size=1,prob=0.5)
  # if coin is heads add a common pattern to that row
  if(coinFlip){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,3),each=5)
  }
}
```

---
class: inverse
## Matrix Data: add a pattern

```{r, fig.width = 8, fig.height = 6}
pheatmap(dataMatrix,cluster_rows=FALSE,cluster_cols=FALSE)
```

---
class: inverse
## Matrix Data: add a pattern

```{r, fig.width = 8, fig.height = 6}
pheatmap(dataMatrix)
```




---
class: inverse
## Row and column patterns

```{r, echo = FALSE}
par(mfrow=c(1,3))
```

```{r, fig.width = 15, fig.height = 5}
hh <- hclust(dist(dataMatrix)); dataMatrixOrdered <- dataMatrix[hh$order,]; par(mfrow=c(1,3))
pal = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1], col = pal)
plot(rowMeans(dataMatrixOrdered), 40:1, xlab="Row Mean", ylab="Row", pch=19)
plot(colMeans(dataMatrixOrdered), xlab="Column", ylab="Column Mean", pch=19)
```

```{r, echo = FALSE}
par(mfrow=c(1,1))
```

---
class: inverse
## A simple plotting function (for mildly-better plots)

```{r}
myplot = function(x, y, type = "p", ...) {
  if (missing(y)) {
    plot(x, ..., type = type, cex.axis = 2, cex.lab = 2, pch = 19)
  } else {
    plot(x, y, ..., type = type, cex.axis = 2, cex.lab = 2, pch = 19)
  }}
```

```{r echo = FALSE, fig.width = 15, fig.height = 5}
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1], col = pal)
myplot(rowMeans(dataMatrixOrdered), 40:1, xlab="Row Mean", ylab="Row")
myplot(colMeans(dataMatrixOrdered), xlab="Column", ylab="Column Mean")
```

---
class: inverse

## Related problems

.super[

You have multivariate variables $X_{1},...,X_{n}$ 
* Find a new set of multivariate variables that are uncorrelated and explain as much variance as possible.
<br><br>
* If you put all the variables together in one matrix, find the best matrix created with fewer variables (lower rank) that explains the original data.

The first goal is statistical and the second goal is data compression.
]

---
class: inverse

## Related solutions

.super[
SVD

* If X is a matrix with each variable in a column and each observation in a row then the SVD is a "matrix decomposition"

$$
X = UDV^{T}
$$

where the columns of U are orthogonal (left singular vectors), the columns of V are orthogonal (right singular vectors) and D is a diagonal matrix (singular values).

PCA

* The principal components are equal to the right singular values if you first scale (subtract the mean, divide by the standard deviation) the variables.
]

```{r pca_math, results='asis', echo = FALSE}
bg_slide("pca_math", 
  folder = folder,
  size = "70%",
  title = "PCA",
  add_opts = "class: inverse")
```

```{r pca_twovar, results='asis', echo = FALSE}
bg_slide("pca_twovar", 
  folder = folder,
  size = "45%",
  title = "PCA with two variables",
  positions = "bottom",
  add_opts = "class: inverse")
```

---
class: inverse
## Example in two dimensions

```{r, echo = FALSE}
par(mfrow=c(1,3))
```

```{r prcomp}
x = rnorm(100); y = rnorm(100, mean=x);dat = cbind(x,y); 
pca1 = prcomp(dat)
```

```{r, echo = FALSE, fig.width = 12, fig.height = 6}
par(mfrow=c(1,2))
myplot(dat)
myplot(x-mean(x),y-mean(y))
abline(-pca1$rotation[,1])
abline(-pca1$rotation[,2])
```

```{r, echo = FALSE}
par(mfrow=c(1,1))
```


```{r svd_math, results='asis', echo = FALSE}
bg_slide("svd_math", 
  folder = folder,
  size = "80%",
  title = "SVD math!",
  add_opts = "class: inverse")
```


```{r svd_qr, results='asis', echo = FALSE}
bg_slide("svd_qr", 
  folder = folder,
  size = "80%",
  title = "SVD best q rank approximation",
  add_opts = "class: inverse")
```

```{r svd_math2, results='asis', echo = FALSE}
bg_slide("svd_math2", 
  folder = folder,
  size = "70%",
  title = "More facts...",
  add_opts = "class: inverse")
```


```{r svd_picture, results='asis', echo = FALSE}
bg_slide("svd_picture", 
  folder = folder,
  size = "70%",
  title = "SVD in pictures",
  add_opts = "class: inverse")
```

.footnote[https://dl.dropboxusercontent.com/u/7710864/jhsph753/lectures/vadim.pdf]


```{r svd_cat, results='asis', echo = FALSE}
bg_slide("svd_cat", 
  folder = folder,
  size = "90%",
  title = "SVD and residuals",
  add_opts = "class: inverse")
```
 
```{r svd_compression, results='asis', echo = FALSE}
bg_slide("svd_compression", 
  folder = folder,
  size = "80%",
  title = "SVD compression",
  add_opts = "class: inverse")
```

.footnote[https://dl.dropboxusercontent.com/u/7710864/jhsph753/lectures/vadim.pdf]

```{r svd_covar, results='asis', echo = FALSE}
bg_slide("svd_covar", 
  folder = folder,
  size = "80%",
  title = "An important fact",
  add_opts = "class: inverse")
```


```{r svd_genome3, results='asis', echo = FALSE}
bg_slide("svd_genome3", 
  folder = folder,
  size = "55%",
  title = "A widely used trick",
  positions = "bottom",
  add_opts = "class: inverse")
```


---
class: inverse
## Back to our example



```{r svd, eval = FALSE}
svd1 <- svd(scale(dataMatrixOrdered))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
myplot(svd1$u[,1],40:1,,xlab="Row",ylab="First left singular vector")
myplot(svd1$v[,1],xlab="Column",ylab="First right singular vector")
```

```{r svd_show, echo = FALSE, fig.width = 15, fig.height = 5.5}
par(mfrow=c(1,3))
svd1 <- svd(scale(dataMatrixOrdered))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
myplot(svd1$u[,1],40:1,,xlab="Row",ylab="First left singular vector")
myplot(svd1$v[,1],xlab="Column",ylab="First right singular vector")
```

```{r, echo = FALSE}
par(mfrow=c(1,2))
```

---
class: inverse
## Components of SVD - variance explained

```{r, fig.width = 12, fig.height = 6}
par(mfrow=c(1,2))
myplot(svd1$d,xlab="Column",ylab="Singular value")
myplot(svd1$d^2/sum(svd1$d^2),xlab="Column",ylab="Prop. of variance explained")
```



---
class: inverse
## Relationship to PCs

```{r, fig.width = 8, fig.height = 6}
svd1 <- svd(scale(dataMatrixOrdered)); pca1 <- prcomp(dataMatrixOrdered,scale=TRUE)
myplot(pca1$rotation[,1],svd1$v[,1], xlab="Principal Component 1",ylab="Right Singular Vector 1")
abline(c(0,1))
```

---
class: inverse
## Back to variance explained


```{r}
constantMatrix <- dataMatrixOrdered*0
for(i in 1:nrow(dataMatrixOrdered)) constantMatrix[i,] <- rep(c(0,1),each=5)
svd1 <- svd(constantMatrix)
```

```{r, echo = FALSE, fig.width = 15, fig.height = 5}
par(mfrow=c(1,3))
image(t(constantMatrix)[,nrow(constantMatrix):1],
col=pal)
myplot(svd1$d,xlab="Column",ylab="Singular value")
myplot(svd1$d^2/sum(svd1$d^2),xlab="Column", ylab="Prop. of variance explained")
```

---
class: inverse
## What if we add a second pattern?

```{r}
set.seed(678910)
for(i in 1:40){
  # flip a coin
  coinFlip1 <- rbinom(1,size=1,prob=0.5)
  coinFlip2 <- rbinom(1,size=1,prob=0.5)
  # if coin is heads add a common pattern to that row
  if(coinFlip1){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5),each=5)
  }
  if(coinFlip2){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5),5)
  }
}
hh <- hclust(dist(dataMatrix)); dataMatrixOrdered <- dataMatrix[hh$order,]
```

---
class: inverse
## What if we add a second pattern?

```{r, fig.width = 15, fig.height = 5}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1],col=pal)
myplot(rep(c(0,1),each=5),xlab="Column",ylab="Pattern 1")
myplot(rep(c(0,1),5),xlab="Column",ylab="Pattern 2")
```

---
class: inverse
## v and row patterns

```{r, fig.width = 15, fig.height = 5}
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1],col=pal)
plot(svd2$v[,1],xlab="Column",ylab="First right singular vector")
plot(svd2$v[,2],xlab="Column",ylab="Second right singular vector")
```

---
class: inverse
## A pathological case

```{r, fig.width = 10, fig.height = 5}
dataMatrix <- matrix(rnorm(400),nrow=40); dataMatrix[1,] = 50*rep(c(0,1),each=5)
ss = svd(dataMatrix - rowMeans(dataMatrix))
```
.left-column-equal[
```{r, fig.width = 6, fig.height = 5}
myplot(ss$v[,1]); 
```
]
.right-column-equal[
```{r, fig.width = 6, fig.height = 5}
pheatmap(dataMatrix)
```
]
---
class: inverse
## Missing values

```{r, fig.width = 12, fig.height = 6}
dataMatrix2 <- dataMatrixOrdered

## Randomly insert some missing data
dataMatrix2[sample(1:100,size=40,replace=FALSE)] <- NA

# fails because scale doesn't like missing data!
# svd_miss <- svd(scale(dataMatrix2)) 
```

---
class: inverse
## Missing values

```{r, fig.width = 12, fig.height = 6}
library(impute)  ## Available from http://bioconductor.org
data_imputed <- impute.knn(dataMatrix2)$data
svd_non_miss <- svd(scale(data_imputed)); svd_full <- svd(scale(dataMatrixOrdered))
```
```{r echo = FALSE, fig.width = 12, fig.height = 6}
par(mfrow=c(1,2)); 
ylim = range(c(svd_full$v[,1], svd_non_miss$v[,1]))
myplot(svd_full$v[,1], main = "Full data", ylim = ylim)
myplot(svd_non_miss$v[,1], main = "Imputed data", ylim = ylim)
```

```{r digits, results='asis', echo = FALSE}
bg_slide("digits", 
  folder = folder,
  size = "50%",
  title = "Digits example",
  add_opts = "class: inverse")
```

.footnote[http://statweb.stanford.edu/~tibs/ElemStatLearn/]


```{r threes, results='asis', echo = FALSE}
bg_slide("threes", 
  folder = folder,
  size = "70%",
  title = "Validation in threes",
  add_opts = "class: inverse")
```

.footnote[http://statweb.stanford.edu/~tibs/ElemStatLearn/]

```{r pagerank2, results='asis', echo = FALSE}
bg_slide("pagerank2", 
  folder = folder,
  size = "70%",
  title = "Pagerank",
  add_opts = "class: inverse")
```

.footnote[http://statweb.stanford.edu/~tibs/ElemStatLearn/]



```{r pagerank_math, results='asis', echo = FALSE}
bg_slide("pagerank_math", 
  folder = folder,
  size = "70%",
  title = "Random surfer model",
  add_opts = "class: inverse")
```

```{r pagerank_eigen, results='asis', echo = FALSE}
bg_slide("pagerank_eigen", 
  folder = folder,
  size = "65%",
  title = "Pagerank as eigenvector problem",
  add_opts = "class: inverse")
```

```{r pagerank_binary, results='asis', echo = FALSE}
bg_slide("pagerank_binary", 
  folder = folder,
  size = "80%",
  title = "Example",
  add_opts = "class: inverse")
```


---
class: inverse
## Calculate Page Ranks

```{r pagerank}
L = matrix(
  c(0, 0, 1, 0, 1, 0, 0, 0, 1, 1, 0, 1, 0, 0, 0, 0),
  byrow = TRUE,
  nrow = 4)
cc = colSums(L)
d = 0.85
e = rep(1, 4)
N = 4
A = (1 - d) * e %*% t(e) + d * L %*% solve(diag(cc))
Re(eigen(A)$values)
phat = Re(eigen(A)$vectors[1,])
phat = (N * phat / c( t(e) %*% phat))
phat
```

---
class: inverse
# An example of interpretable pcs

```{r genes-geo, results='asis', echo = FALSE}
bg_slide("genes-geography", 
  folder = folder,
  size = "80%",
  title = "Genetics PCs",
  add_opts = "class: inverse")
```

---
class: inverse

## Dimension Reduction

.super[
- PCA is a highly-useful (and used tool)
<br><br>
- Many use variance explained threshold for dimension reduction
<br><br>
- Also an orthogonalizing tool
<br><br>
- Maximum Rank of matrix X is always $\min(n, p)$
<br><br>
- Use the lower-dimensional matrix for big data
<br><br>
]

