---
title: "Machine Learning"
subtitle: "JHU Data Science"
author: "www.jtleek.com/advdatasci"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "../additional.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    keep_md: true
---
class: inverse, middle, center

# Prediction 

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "")
source("../slide_functions.R")
folder = "../imgs/ml/"
knitr::opts_chunk$set(
  echo = TRUE,
  prompt = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = ""
  )
library(RColorBrewer)
library(RSkittleBrewer)
library(ggplot2)
library(pheatmap)
theme_big = theme(text = element_text(size = 20))
```

---
class: inverse

## Other names for prediction 

.huge[
* Prediction -> statisticians
* Statistical learning -> statisticians
* Machine learning -> computer scientists
* Forecasting -> atmospheric scientists/bankers
* Artificial intelligence -> the popular press
]

```{r, results='asis', echo = FALSE}
bg_slide("selfcar", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```


```{r, results='asis', echo = FALSE}
bg_slide("whetlab", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```


```{r, results='asis', echo = FALSE}
bg_slide("self_drive", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```


```{r, results='asis', echo = FALSE}
bg_slide("blackbox", 
  folder = folder,
  size = "60%",
  add_opts = "class: inverse")
```


.footnote[https://twitter.com/notajf/status/795717253505413122]

---
class: inverse

## Key idea: know why x predicts y

.huge[
>The combination of some data and an aching desire for an answer does not ensure that a reasonable answer can be extracted from a given body of data. <br><br>-John Tukey
]

```{r, results='asis', echo = FALSE}
bg_slide("nobel", 
  folder = folder,
  size = "70%",
  add_opts = "class: inverse")
```
.footnote[http://www.nejm.org/doi/full/10.1056/NEJMon1211064]


```{r, results='asis', echo = FALSE}
bg_slide("prob", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("study_design", 
  folder = folder,
  size = "40%",
  title = "Study design",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("overfitting", 
  folder = folder,
  size = "80%",
  titles = "Overfitting",
  add_opts = "class: inverse")
```

.footnote[http://blog.fliptop.com/blog/2015/03/02/bias-variance-and-overfitting-machine-learning-overview/]


```{r, results='asis', echo = FALSE}
bg_slide("training_data", 
  folder = folder,
  size = "65%",
  titles = "Training data matters a lot",
  add_opts = "class: inverse")
```

.footnote[http://www.google.org/flutrends/]

```{r, results='asis', echo = FALSE}
bg_slide("failed", 
  folder = folder,
  size = "80%",
  titles = "Changes in population -> failed predictions",
  add_opts = "class: inverse")
```

.footnote[http://gking.harvard.edu/files/gking/files/0314policyforumff.pdf]

```{r, results='asis', echo = FALSE}
bg_slide("machine_bias", 
  folder = folder,
  size = "80%",
  titles = "Prediction != unbiased",
  add_opts = "class: inverse")
```


.footnote[https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing]

```{r, results='asis', echo = FALSE}
bg_slide("science", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("science_retract", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("prediction_vs_assoc", 
  folder = folder,
  size = "80%",
  titles = "Prediction vs. association",
  add_opts = "class: inverse")
```

.footnote[Slide courtesy Ingo Ruczinski]

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

.footnote[http://en.wikipedia.org/wiki/Sensitivity_and_specificity]

---
class: inverse

## Most common error measures

.super[
1. Mean squared error (or root mean squared error)
    - Continuous data, sensitive to outliers
2. Median absolute deviation
    - Continuous data, often more robust
3. Sensitivity (recall)
    - If you want few missed positives
4.  Specificity
    - If you want few negatives called positives
5. Accuracy
    - Weights false positives/negatives equally
6. Concordance
    - One example is kappa
7.  Predictive value of a positive (precision)
    - When you are screeing and prevelance is low
]


```{r, results='asis', echo = FALSE}
bg_slide("roc", 
  folder = folder,
  size = "50%",
  title = "ROC",
  add_opts = "class: inverse")
```



```{r, results='asis', echo = FALSE}
bg_slide("precision_recall", 
  folder = folder,
  size = "70%",
  add_opts = "class: inverse")
```





```{r, results='asis', echo = FALSE}
bg_slide("recall_curve", 
  folder = folder,
  size = "45%",
  title = "Precision/Recall",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("measure_good", 
  folder = folder,
  size = "50%",
  title = 'Measuring "good"',
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec2", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec3", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec4", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("ppv", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec5", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("sens_spec6", 
  folder = folder,
  size = "80%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("depends", 
  folder = folder,
  size = "75%",
  title = "Depends on your problem",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("prostate", 
  folder = folder,
  size = "65%",
  title = "Depends on your problem",
  add_opts = "class: inverse")
```


```{r, results='asis', echo = FALSE}
bg_slide("mamma", 
  folder = folder,
  size = "90%",
  add_opts = "class: inverse")
```

---
class: inverse, middle, center

.sixer[
89% sensitivity<br>
42% specificity<br>
65% accuracy<br>
]

```{r, results='asis', echo = FALSE}
bg_slide("tsp", 
  folder = folder,
  size = "130%",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("spam", 
  folder = folder,
  size = "50%",
  titles = "Steps in the process\n question -> input data -> features -> algorithm -> parameters -> evaluation",
  positions = "bottom",
  add_opts = "class: inverse")
```

```{r, results='asis', echo = FALSE}
bg_slide("time_series", 
  folder = folder,
  size = "80%",
  titles = "Time series data",
  positions = "bottom",
  add_opts = "class: inverse")
```

---
class: inverse

## What is different?

.super[
* Data are dependent over time
* Specific pattern types
    * Trends - long term increase or decrease
    * Seasonal patterns - patterns related to time of week, month, year, etc.
    * Cycles - patterns that rise and fall periodically
* Subsampling into training/test is more complicated
* Similar issues arise in spatial data
    * Dependency between nearby observations
    * Location specific effects
* Typically goal is to predict one or more observations into the future.
* All standard predictions can be used (with caution!)
]

```{r, results='asis', echo = FALSE}
bg_slide("spurious", 
  folder = folder,
  size = "70%",
  titles = "Beware spurious correlations",
  add_opts = "class: inverse")
```

.footnote[http://www.google.com/trends/correlate,<br>
http://www.newscientist.com/blogs/onepercent/2011/05/google-correlate-passes-our-we.html]


```{r, results='asis', echo = FALSE}
bg_slide("geog", 
  folder = folder,
  size = "45%",
  positions = "bottom",
  titles = "Also common in geographic analysis",
  add_opts = "class: inverse")
```

.footnote[http://xkcd.com/1138/]

```{r, results='asis', echo = FALSE}
bg_slide("test_sets", 
  folder = folder,
  size = "80%",
  titles = "Careful with picking your test sets",
  add_opts = "class: inverse")
```

.footnote[http://waldronlab.org/wp-content/uploads/2014/11/Waldron-et-al.-2014-Comparative-meta-analysis-of-prognostic-gene-signatures-for-late-stage-ovarian-cancer.pdf]

```{r, results='asis', echo = FALSE}
bg_slide("long_con", 
  folder = folder,
  size = "45%",
  title = "An internet scam that is relevant to you",
  add_opts = "class: inverse")
```

.footnote[https://medium.com/message/how-to-always-be-right-on-the-internet-delete-your-mistakes-519a595da2f5#.lutyyhyex]
