---
title: "Make and JHPCE"
subtitle: "JHU Data Science"
author: "www.jtleek.com/advdatasci"
output: 
  ioslides_presentation:
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is `make`?

## Make

GNU Make is a tool which controls the generation of executables and other non-source files of a program from the program's source files.

- set of instructions
- builds "exectutables", could be:
    - a pdf
    - a markdown document
- can incorporate dependencies
    
Karl Broman and Make: http://kbroman.org/minimal_make/


## How do use `make`?

- Have a `Makefile` (spelled `makefile` or `Makefile`)

```make
target: dependencies
TAB instruction 1
TAB instruction 2
```

## What is a `target`? Example `Makefile`

- A target is something you want to build.  Here is an example:

```make
all: index.html index.R 
    
index.html: index.Rmd 
	Rscript -e "rmarkdown::render('index.Rmd')"

index.R: index.Rmd
	Rscript -e "knitr::purl('index.Rmd')"

clean: 
	rm index.html index.R
```

- `all` is special


## How do you run this?

In the terminal:

```bash
make # will run all
make all
make clean
make index.html
make index.R
```

## So it's like a script, so what?

- Make decides whether a target needs to be regenerated by **comparing file modification times.** (Wikipedia)

- **This solves the problem of avoiding the building of files which are already up to date, but it fails when a file changes but its modification time stays in the past.**

- Almost like caching, can save you a lot of time


## Some catches

- Make **requires tabs, not spaces**.  Make sure your text editor for your `Makefile` is configured properly.

- Changing directories need to be done **the same line**:

```make
index.html: R/somefile.R
	cd R
	Rscript somefile.R
```

will not work.

## Single line executions

```make
index.html: R/somefile.R
	cd R; Rscript somefile.R
```

If you use the backslash, then it's seen as a continuance of the last line.  
Note, the next line still needs to be indented for `make` to know it's still in the same target.

```make
index.html: R/somefile.R 
	cd R; \
	Rscript somefile.R
```



## Single line executions

If statements can be broken up but again need to be with continuation lines:

```make
index.html: R/somefile.R
 	if [ -e ${name}.aux ]; \
	then \
	rm ${name}.aux; \
	fi;
```

**If you have spaces after the `\`, then it doesn't work!**


## Splitting up dependencies

Dependencies can be broken up into multiple lines:

```make
target: targetA targetB \
targetC targetD
  recipe
```

or defined on multiple lines

```make
target: targetA targetB
target: targetC targetD

target:
  recipe
```


# JHPCE computing and R

## Setting up a passwordless login for your local machine

https://jhpce.jhu.edu/knowledge-base/authentication/ssh-key-setup/

# How do you submit an R script job to the cluster

## Submitting an R script job

- `qsub` - submits a job the **q**ueue
- By default, it assumes you're submitting a shell script that has one of 2 lines on there:

```{bash, eval = FALSE}
Rscript script.R
R CMD batch script.R
```

**What's the difference?** (Someone say something here...)

## Submitting an R script job

- Not great if you want to just run an R script without creating a shell script
- If you have a `script.R` file, you can execute this a few ways:
    - Put `#!/usr/bin/env RScript` as the first line of your file
    - Use [R functions to help](../R_functions.sh).
- Rscript doesn't load the `methods` package and allows for command-line arguments

# What is an array job?

## The `Sys.getenv` function 

- `Sys.getenv` stands for get System environmental variables
- These are in `.bash_profile` or `.bashrc`, and say something like:
    - `export VARIABLE=something`

On JHPCE cluster, `SGE_TASK_ID` is the task identifier for a job.  It is set when you use the `-t` option in `qsub`:

```bash
qsub script.R -t 1-100
```


## Task IDs and `expand.grid`

- `expand.grid`: Creates a data frame from all combinations of the supplied vectors or factors.

```{r}
eg = expand.grid(param1 = 1:10, param2 = c(3, 5, 6))
head(eg)
sim_num = Sys.getenv("SGE_TASK_ID")
sim_num = as.numeric(sim_num) # Sys.getenv returns char
if (is.na(sim_num)) {
  sim_num = 1
}
this_sim = eg[sim_num,]
param1 = this_sim$param1
param2 = this_sim$param2
```






