---
title: "Data Knick Knacks and Strings"
author: "John Muschelli"
date: "September 21, 2016"
output: 
  ioslides_presentation:
    widescreen: yes
    css: muschelli.css
---


```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(comment = "")
library(plyr)
library(dplyr)
```


## A funny quote

> "Herein lies the dirty secret about most data scientists' work -- it's more data munging than deep learning. The best minds of my generation are deleting commas from log files, and that makes me sad. A Ph.D. is a terrible thing to waste."

http://adage.com/article/digitalnext/dear-madison-avenue-set-data-scientists-free/298676/

## Raw versus processed data

<div class="container">
<div class="left-half">
__Raw data__

* The original source of the data
* Often hard to use for data analyses
* Data analysis _includes_ processing
* Raw data may only need to be processed once

[http://en.wikipedia.org/wiki/Raw_data](http://en.wikipedia.org/wiki/Raw_data)
</div>
<div class="right-half">
__Processed data__

* Data that is ready for analysis
* Processing can include merging, subsetting, transforming, etc.
* There may be standards for processing
* All steps should be recorded 

[http://en.wikipedia.org/wiki/Computer_data_processing](http://en.wikipedia.org/wiki/Computer_data_processing)
</div>
</div>


## Tidy data

1. Each variable you measure should be in one column
2. Each different observation of that variable should be in a different row
3. There should be one table for each "kind" of variable
4. If you have multiple tables, they should include a column in the table that allows them to be linked

_Some other important tips_

* Include a row at the top of each file with variable names. 
* Make variable names human readable AgeAtDiagnosis instead of AgeDx
* In general data should be saved in one file per table.

[https://github.com/jtleek/datasharing](https://github.com/jtleek/datasharing)
[Tidy data](http://www.jstatsoft.org/v59/i10/paper)


## The four things you should have

1. The raw data.
2. A tidy data set
3. A code book describing each variable and its values in the tidy data set.
4. An explicit and exact recipe you used to go from 1 -> 2,3.


## Raw data


<img class=center src=./dsp1.png height=450>


## Tidy data


<img class=center src=./dsp2.png height=450>


## Code book


<img class=center src=./dsp3.png height=450>



## Recipe


<img class=center src=./dsp4.png height=450>



## Recipe ok


<img class=center src=./dsp5.png height=450>

## Recipe baadddd


<img class=center src=./dsp6.png height=450>


## This is hard

<img class=center src=./broman-repro-hard.png height=450>

https://kbroman.wordpress.com/2015/09/09/reproducibility-is-hard/

# Project Structure

- RStudio Projects
- Project template: http://projecttemplate.net/getting_started.html
- File structure:
    - raw_data
    - processed_data
    - figures
    - programs

Either way, it's good to have a good handle on how to deal with files and 
manipulate them.


## Quick Aside: File Manipulations

Let's also look at `?files` and `dir.exists`

```{r}
if (!dir.exists("data")) {
  dir.create("data")
}
file.exists("data")
dir.exists("data")
file.exists("Data_Cleaning.Rmd")
dir.exists("Data_Cleaning.Rmd")
```

## Quick Aside: File Manipulations

```{r}
x = list.files("data")
x
x = list.files(pattern = "[.][R|r]md")
file.create("test.rmd")
list.files(pattern = "[.][R|r]md")
file.remove("test.rmd")
list.files(pattern = "[.][R|r]md")
```



## System Calls 

Sometimes you just want to use the `mkdir` command or something in your shell:
```{r, eval = FALSE}
system("COMMAND")
```

# Reading in Data

## Most common file format


<img class=center src=./excel.png height=450>

http://office.microsoft.com/en-us/excel/


## 1,000 genomes example

We want to download a dataset, read it in, but not store it on my hard drive.
We will use `tempfile` for that (see `tempdir()`)

```{r, cache=TRUE}
fileUrl = paste0("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp",
                 "/technical/working/20130606_sample_info/20130606_sample_info.xlsx")
destfile = tempfile(fileext = ".xlsx")
print(destfile)
file.exists(destfile)
```


## Downloading data: downloader package

Save `dateDownloaded` with the object you read `destfile` into:

```{r cache = TRUE}
library(downloader)
download(fileUrl, destfile = destfile)
dateDownloaded <- date()
dateDownloaded
file.exists(destfile)
```

## Reading in data with Excel, different sheets

```{r cache = TRUE}
library(readxl)
# Have to skip one row because 
# there is an extra header
kg = read_excel(destfile,
                sheet = 4, 
                skip = 1)
kg2 = read_excel(destfile,
                sheet = 4, 
                skip = 1)
all.equal(as.data.frame(kg), as.data.frame(kg2))
all.equal(kg, kg2)
```


## Reading data from text files

<div class="container">
<div class="left-half">
___Normal Way___

```{r, eval=F}
read.table
read.csv
```
</div>
<div class="right-half">
___Newer Way (faster)___
```{r, eval=F}
read_csv 
fread
```
</div>
</div>

## Let's make a big(ish) file

```{r}
library(readr)
dat = matrix(rnorm(1000*1000),
             nrow = 1000)
outfile = tempfile(fileext = ".csv")
write.csv(dat, file = outfile)
file.exists(outfile)
```

## Quick comparison: read_csv is faster

<div class="container">
<div class="left-half">
```{r}
system.time({x = read.csv(outfile)})
class(x)
```
</div>
<div class="right-half">
```{r}
system.time({x = read_csv(outfile)})
class(x)
```
</div>
</div>

## Quick comparison: fread is faster

```{r}
library(data.table)
```

## Quick comparison: fread is faster

```{r}
system.time({x = fread(outfile)})
class(x)
```


# A little data cleaning 

## Finding Missing data

Each missing data type has a function that returns `TRUE` if the data is missing:

* `NA` - `is.na`
* `NaN` - `is.nan`
* `Inf` and `-Inf` - `is.infinite`
* `is.finite` returns `FALSE` for all missing data and `TRUE` for non-missing
* `complete.cases` on a `data.frame`/`matrix` returns `TRUE` if all values in that row of the object are not missing.

## Missing Data with Logicals

One important aspect (esp with subsetting) is that logical operations return `NA` for `NA` values.  Think about it, the data could be `> 2` or not we don't know, 
so `R` says there is no `TRUE` or `FALSE`, so that is missing:
```{r}
x = c(0, NA, 2, 3, 4)
x > 2
x[x > 2]
```


## Missing Data with Logicals

What about seeing if a value is equal to multiple values?  You can do `(x == 1 | x == 2) & !is.na(x)`, but that is not efficient.  Use `%in%`

```{r}
(x == 0 | x == 2) # has NA
(x == 0 | x == 2) & !is.na(x) # No NA
x %in% c(0, 2) # NEVER has NA and returns logical
```


# Tables and Tabulations

## Creating One-way Tables 

Here we will use `table` to make tabulations of the data.   Look at `?table` to see options for missing data:

```{r table}
table(x)
```


From `table` source code (default for `useNA`?): 
```{r table_out}
table %>% head(22) %>% tail(1)
```


## Creating One-way Tables 

You can set `useNA = "always"` to have it always have a column for `NA`
```{r onetab}
table(x, useNA = "ifany")
table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
      useNA = "always")
```


## Creating Two-way Tables 

A two-way table.  If you pass in 2 vectors, `table` creates a 2-dimensional table.

```{r}
tab <- table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
             c(0, 1, 2, 3, 2, 3, 3, 4, 4, 3), 
             useNA = "always")
```

## Finding Row or Column Totals

`margin.table` finds the marginal sums of the table.  `margin` is 1 for rows, 2 for columns in general in `R`.  Here is the column sums of the table:

```{r margin}
margin.table(tab, 2)
```

## Proportion Tables 

`prop.table` finds the marginal proportions of the table. Think of it dividing the table by it's respective marginal totals.  If `margin` not set, divides by overall total.

```{r table2}
prop.table(tab)
prop.table(tab,1)
```

## Proportion Tables 

`margin = 1` means row percentages

```{r table2_again}
prop.table(tab,1)
```

## Download Salary FY2014 Data

From https://data.baltimorecity.gov/City-Government/Baltimore-City-Employee-Salaries-FY2014/2j28-xzd7
http://www.aejaffe.com/winterR_2016/data/Baltimore_City_Employee_Salaries_FY2014.csv

Read the CSV into R `Sal`:

```{r readSal}
Sal = read.csv(
  paste0("http://www.aejaffe.com/winterR_2016/data",
         "/Baltimore_City_Employee_Salaries_FY2014.csv"),
  as.is = TRUE)
```

## Checking for logical conditions 
* `any()` - checks if there are any `TRUE`s
* `all()` - checks if ALL are true

```{r isna}
head(Sal,2)
any(is.na(Sal$Name)) # are there any NAs?
```

# Recoding Variables

## Example of Recoding: base R

For example, let's say gender was coded as Male, M, m, Female, F, f.  Using Excel to find all of these would be a matter of filtering and changing all by hand or using if statements.

In R, you can simply do something like:

```{r, eval = FALSE}
data$gender[data$gender %in% 
              c("Male", "M", "m")] <- "Male"
```

## Example of Recoding with `recode`: `car` package

You can also `recode` a vector:

```{r}
library(car, quietly = TRUE)

car::recode(x, "c('m', 'M', 'male') = 'Male';
            c('f', 'F', 'female') = 'Female';")
```

## Example of Recoding with `revalue`: `plyr`

You can `revalue` a vector with the `revalue` command

```{r}
library(plyr)
x = rep(c("Male", "M", "m", "f", "Female", "female" ), 
        each = 3)
plyr::revalue(x, c("M" = "Male", "m" = "Male",
                   "f" = "Female", "female" = "Female"))
```

## Example of Cleaning: more complicated

Sometimes though, it's not so simple.  That's where functions that find patterns come in very useful.

```{r gender, echo=TRUE}
set.seed(4) # random sample below - make sure same every time
gender <- sample(c("Male", "mAle", "MaLe", "M", "MALE", 
                   "Ma", "FeMAle", "F", "Woman", "Man", "Fm", "FEMALE"), 
                 1000, replace = TRUE)
```

```{r gentab}
table(gender)
```


## Example of Cleaning: use dput

`dput` "Writes an ASCII text representation of an R object to a file or connection, or uses one to recreate the object."

```{r gender_dput, echo=TRUE}
dput(unique(gender))
# Copy this to your clipboard
# stopifnot( all(gender %in% CLIPBOARD_PASTE))
```


# String functions

## Pasting strings with `paste` and `paste0`

Paste can be very useful for joining vectors together:

```{r Paste}
paste("Visit", 1:5, sep = "_")
paste("Visit", 1:5, sep = "_", collapse = " ")
paste("To", "is going be the ", "we go to the store!", sep = "day ")
# and paste0 can be even simpler see ?paste0 
paste0("Visit",1:5)
```

## Paste Depicting How Collapse Works

```{r Paste2}
paste(1:5)
paste(1:5, collapse = " ")
```

## Useful String Functions

Useful String functions

* `toupper()`, `tolower()` - uppercase or lowercase your data:
* `str_trim()` (in the `stringr` package) or `trimws` in base 
- will trim whitespace
* `nchar` - get the number of characters in a string
* `paste()` - paste strings together with a space 
* `paste0` - paste strings together with no space as default


## The `stringr` package

Like `dplyr`, the `stringr` package:

* Makes some things more intuitive
* Is different than base R
* Is used on forums for answers
* Has a standard format for most functions
* the first argument is a string like first argument is a `data.frame` in `dplyr`



## Splitting/Find/Replace and Regular Expressions

* R can do much more than find exact matches for a whole string
* Like Perl and other languages, it can use regular expressions.
* What are regular expressions?
* Ways to search for specific strings 
* Can be very complicated or simple
* Highly Useful - think "Find" on steroids


## A bit on Regular Expressions

* http://www.regular-expressions.info/reference.html
* They can use to match a large number of strings in one statement
* `.` matches any single character
* `*` means repeat as many (even if 0) more times the last character
* `?` makes the last thing optional
* `^` matches start of vector `^a` - starts with "a"
* `$` matches end of vector `b$` - ends with "b"

# Splitting Strings

## Substringing

Very similar:

Base R

* `substr(x, start, stop)` - substrings from position start to position stop
* `strsplit(x, split)` - splits strings up - returns list!

`stringr`

* `str_sub(x, start, end)` - substrings from position start to position end
* `str_split(string, pattern)` - splits strings up - returns list!


## Splitting String: base R

In base R, `strsplit` splits a vector on a string into a list
```{r strsplit}
x <- c("I really", "like writing", "R code programs")
y <- strsplit(x, split = " ") # returns a list
y
```

## Splitting String: `stringr`

`stringr::str_split` do the same thing:

```{r str_split}
library(stringr)
y2 <- str_split(x, " ") # returns a list
y2
```

## Using a fixed expression

One example case is when you want to split on a period "`.`".  In regular expressions `.` means **ANY** character, so

```{r}
str_split("I.like.strings", ".")
str_split("I.like.strings", fixed("."))
```

## Let's extract from `y`

```{r stsplit2}
suppressPackageStartupMessages(library(dplyr)) # must be loaded AFTER plyr
y[[2]]
sapply(y, dplyr::first) # on the fly
sapply(y, nth, 2) # on the fly
sapply(y, last) # on the fly
```


## 'Find' functions: base R

`grep`: `grep`, `grepl`, `regexpr` and `gregexpr` search for matches to argument pattern within each element of a character vector: they differ in the format of and amount of detail in the results. 

`grep(pattern, x, fixed=FALSE)`, where:

* pattern = character string containing a regular expression to be matched in the given character vector.

* x = a character vector where matches are sought, or an object which can be coerced by as.character to a character vector.

* If fixed=TRUE, it will do exact matching for the phrase anywhere in the vector (regular find)


## 'Find' functions: `stringr`

`str_detect`, `str_subset`, `str_replace`, and `str_replace_all` search for matches to argument pattern within each element of a character vector: they differ in the format of and amount of detail in the results. 

* `str_detect` - returns `TRUE` if `pattern` is found
* `str_subset` - returns only the strings which pattern were detected
* convenient wrapper around `x[str_detect(x, pattern)]`
* `str_extract` - returns only strings which pattern were detected, but ONLY the pattern
* `str_replace` - replaces `pattern` with `replacement` the first time
* `str_replace_all` - replaces `pattern` with `replacement` as many times matched

## 'Find' functions: stringr compared to base R

Base R does not use these functions.  Here is a "translator" of the `stringr` function to base R functions

* `str_detect` - similar to `grepl` (return logical)
* `grep(value = FALSE)` is similar to `which(str_detect())`
* `str_subset` - similar to `grep(value = TRUE)` - return value of matched
* `str_replace` - similar to `sub` - replace one time 
* `str_replace_all` - similar to `gsub` - replace many times

## Let's look at modifier for `stringr`

`?modifiers`

* `fixed` - match everything exactly
* `regexp` - default - uses **reg**ular **exp**ressions
* `ignore_case` is an option to not have to use `tolower`


## Important Comparisons

Base R:

* Argument order is `(pattern, x)`
* Uses option `(fixed = TRUE)`

`stringr`

* Argument order is `(string, pattern)` aka `(x, pattern)`
* Uses function `fixed(pattern)`



## 'Find' functions: finding values, `stringr` and `dplyr` 

```{r ggrep}
str_subset(Sal$Name, "Rawlings")
Sal %>% filter(str_detect(Name, "Rawlings"))
```

## Showing differnce in `str_extract`

`str_extract` extracts just the matched string

```{r ggrep2}
ss = str_extract(Sal$Name, "Rawling")
head(ss)
ss[ !is.na(ss)]
```

## Showing differnce in `str_extract` and  `str_extract_all`

`str_extract_all` extracts all the matched strings (we're matching a digit)

```{r }
head(str_extract(Sal$AgencyID, "\\d"))
head(str_extract_all(Sal$AgencyID, "\\d"), 2)
```


## Using Regular Expressions: `stringr`
```{r grepstar_stringr}
head(str_subset( Sal$Name, "^Payne.*"), 3)
```

```{r grepstar2_stringr}
head(str_subset( Sal$Name, "Leonard.?S"))
head(str_subset( Sal$Name, "Spence.*C.*"))
```

## Replace

Let's say we wanted to sort the data set by Annual Salary:

```{r classSal}
class(Sal$AnnualSalary)
```

```{r orderstring}
sort(c("1", "2", "10")) #  not sort correctly (order simply ranks the data)
order(c("1", "2", "10"))
```

## Replace

So we must change the annual pay into a numeric:
```{r destringSal}
head(Sal$AnnualSalary, 4)
head(as.numeric(Sal$AnnualSalary), 4)
```

R didn't like the `$` so it thought turned them all to `NA`.

`sub()` and `gsub()` can do the replacing part in base R.

## Replacing and subbing 

Now we can replace the `$` with nothing (used `fixed=TRUE` because `$` means ending):

```{r orderSal}
Sal$AnnualSalary <- as.numeric(gsub(pattern = "$", replacement="", 
                                    Sal$AnnualSalary, fixed=TRUE))
Sal <- Sal[order(Sal$AnnualSalary, decreasing=TRUE), ] 
Sal[1:5, c("Name", "AnnualSalary", "JobTitle")]
```

## Replacing and subbing: `stringr` 

We can do the same thing (with 2 piping operations!) in dplyr
```{r orderSal_stringr}
dplyr_sal = Sal
dplyr_sal = dplyr_sal %>% mutate( 
  AnnualSalary = AnnualSalary %>%
    str_replace(
      fixed("$"), 
      "") %>%
    as.numeric) %>%
  arrange(desc(AnnualSalary))
check_Sal = Sal
rownames(check_Sal) = NULL
all.equal(check_Sal, dplyr_sal)
```

